---
title: "Gapminder"
author: "ingridmathilde"
date: 2020-07-22
output:
  github_document:
    toc: true
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)
library(gapminder)
library(ggrepel)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to education people about the state of the world. They promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

__q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
head(gapminder)
glimpse(gapminder)
summary(gapminder)
```

**Observations**:

* **This dataset contains the following information:**
  + **Country**
  + **Continent**
  + **Year**
  + **Life Expentancy**
  + **Population**
  + **GDP per capita**

__q1__ Determine the most and least recent years in the `gapminder` dataset.

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- max(select(gapminder,year))
year_min <- min(select(gapminder, year))
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)
print("Nice!")
```

__q2__ Filter on years matching `year_min`, and make a plot of the GDE per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>% 
  filter(year == year_min) %>% 
  ggplot() +
  geom_boxplot(aes(x = continent, y = gdpPercap)) +
  labs(title = "Continent vs GDP per Capita (full scale)")

gapminder %>% 
  filter(year == year_min) %>% 
  ggplot() +
  geom_boxplot(aes(x = continent, y = gdpPercap), notch = TRUE) +
  coord_cartesian(ylim = c(0,20000))+
  labs(title = "Continent vs GDP per Capita (zoomed y-axis omits one outlier)")

```

***Observations***:

* **Generally, Oceania has a greater GDP per Capita, followed by Europe, then the Americas, then Asia, then Africa.**
* **Per continent there are outliers:**
  + **Africa: 3**
  + **Americas: 3**
  + **Asia : 2**
  + **Europe : 1**
  + **Oceania: 0**
* **The GDP per Capita is the most spread for Europe, followed by Asia, then Americas, then Africa, then Oceania.**

***Difficulties & Approaches***:

* **Because of the greatest outlier for Asia, the scale of the graph made it difficult to understand the data - so I used coord_cartesian to zoom into the data.**

__q3__ You should have found at least three outliers in q2. Identify those outliers (figure out which countries they are).

*Note*: I realize the reasons for

```{r q3-task}
## TASK: Identify the outliers from q2

describe_continent_min <- gapminder %>%
  filter(year == year_min) %>% 
  group_by(continent) %>% 
  mutate(min = min(gdpPercap),
        max = max(gdpPercap),
        median = median(gdpPercap),
        IQRange = IQR(gdpPercap),
        lower = quantile(gdpPercap, 0.25) - 1.5 * IQR(gdpPercap),
        upper = quantile(gdpPercap, 0.75) + 1.5 * IQR(gdpPercap)
            ) %>% 
  ungroup() %>% 
  filter(upper < gdpPercap)

gapminder %>% 
  filter(year == year_min) %>% 
  ggplot(
    aes(x = continent, 
        y = gdpPercap)) +
  geom_boxplot(outlier.color = "red") +
  geom_label_repel(
    data = describe_continent_min, 
    mapping = aes(label = country)) +
  labs(title = "Continent vs GDP per Capita (full scale)")

gapminder %>% 
  filter(year == year_min) %>% 
  ggplot(
    aes(x = continent, 
        y = gdpPercap)) +
  geom_boxplot(notch = TRUE, outlier.color = "red") +
  geom_label_repel(
    data = describe_continent_min, 
    mapping = aes(label = country)) +
  coord_cartesian(ylim = c(0,20000))+
  labs(title = "Continent vs GDP per Capita (zoomed y-axis omits one outlier)")

```

***Observations***:

* **My solution is quick and dirty but hey, it works.**
* **The countries are:**
  + **Kuwait**
  + **Switzerland**
  + **Bahrain**
  + **United States**
  + **Canada**
  + **Venezuela**
  + **South Africa**
  + **Gabon**
  + **Angola**

__q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
describe_continent_max <- gapminder %>%
  filter(year == year_max) %>% 
  group_by(continent) %>% 
  mutate(min = min(gdpPercap),
        max = max(gdpPercap),
        median = median(gdpPercap),
        IQRange = IQR(gdpPercap),
        lower = quantile(gdpPercap, 0.25) - 1.5 * IQR(gdpPercap),
        upper = quantile(gdpPercap, 0.75) + 1.5 * IQR(gdpPercap)
            ) %>% 
  ungroup() %>% 
  filter(upper < gdpPercap)

gapminder %>% 
  filter(year == year_min | year == year_max) %>% 
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot(outlier.color = "red") +
  facet_grid(.~year) +
  labs(title = "Continent vs GDP per Capita (full scale)")
 

gapminder %>% 
  filter(year == year_min | year == year_max) %>% 
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot(outlier.color = "red") +
  geom_label_repel(
    data = describe_continent_min, 
    mapping = aes(label = country)) +
  geom_label_repel(
    data = describe_continent_max, 
    mapping = aes(label = country)) +
  facet_grid(.~year) +
  labs(title = "Continent vs GDP per Capita (full scale)") 
  
```

**Observations**:

* **The GDP across the world has increased from 1952-2007.**
* **The variability of GDP has increased within each continent from 1952-2007.**
* **The countries that continue to be outliers compared to their continent are:**
  + **Gabon**
  + **South Africa**
  + **United States**
  + **Canada**

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

__q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph

gapminder %>%
  filter(year == year_max | year == year_min) %>% 
  ggplot(
    aes(
      x = gdpPercap/mean(gdpPercap), 
      y = lifeExp/mean(lifeExp)
      )
    ) +
  geom_point(aes(color = as.factor(year))) +
  labs(title = "Comparative GDP per capita vs Comparative Life Expectancy", 
       subtitle = "Question 5 - Plot 1 "
       )
  
```
***Observations***
* **In 1952 and in 2007, comparatively higher GDP is associated with comparatively higher life expectancy.**
* **In 1952, the comparaitve disparity in life expectancy and GDP was smaller than those in 2007.**
* **This is a trend in the world, but is this trend true for the 'worst off' countries in the world?**

```{r q5-task2}
## TASK: Your second graph
lowest_gdp <- gapminder %>%
  filter(year == year_min) %>% 
  arrange(gdpPercap)

lowest_life <- gapminder %>%
  filter(year == year_min) %>% 
  arrange(lifeExp)

gapminder %>% 
  filter(country == "Lesotho") %>% 
  ggplot(aes(x = year, y = gdpPercap)) +
  geom_line(color = "blue") +
  geom_point(aes(alpha = lifeExp)) +
  labs(
    title = "Lesotho's (the poorest country in 1952)\n changes in GDP per capita and Life Expectancy",
    subtitle = "Question 5 - Plot 2")

gapminder %>% 
  filter(country == "Afghanistan") %>% 
  ggplot(aes(x = year, y = gdpPercap)) +
  geom_line(color = "red") +
  geom_point(aes(alpha = lifeExp)) +
  labs(
    title = "Afghanistan's (the country with the lowest life expectancy in 1952)\n changes in GDP per capita and Life Expectancy",
    subtitle = "Question 5 - Plot 3")

```

***Observations***

* **The trend previously discuss (higher GDP per capita associated with higher life expectancy) would dictate that as GDP increases, the life expectancy increases.**
* **For the 'worst-off' countries in the world in 1952, there is not a direct correlation between GPD per capita and life expectancy as it changed over time.**
* **Obviously GDP is not the only factor at play - war, famine, health crises can all affect life expectancy. Both [Afghanistan](https://en.wikipedia.org/wiki/History_of_Afghanistan#Contemporary_era_(1973%E2%80%93present)) and [Lesotho](https://en.wikipedia.org/wiki/History_of_Lesotho#Kingdom_of_Lesotho) have experienced armed conflicts, famine, and widespread inaccessibility to healthcare throughout their histories.**


```{r q5-task3}
## TASK: Your third graph

describe_country <- gapminder %>%
  filter(year == year_min | year == year_max & !is.na(gdpPercap) & !is.na(lifeExp)) %>%
  select(-pop) %>% 
  pivot_wider(names_from = year, values_from = c(lifeExp, gdpPercap))

weirdos <- describe_country %>%
  filter(gdpPercap_1952 > gdpPercap_2007 | lifeExp_1952 > lifeExp_2007)
  
  weirdos

describe_country %>% 
  ggplot(
    aes(
      x = gdpPercap_1952, 
      y = lifeExp_1952
      )
  ) +
  geom_segment(
    aes(
      xend = gdpPercap_2007, 
      yend = lifeExp_2007, 
      color = continent
  ), 
    arrow = arrow(length = unit(0.1,"cm"))) +
  geom_label_repel(
    data = filter(weirdos, continent != "Africa"),
    mapping = aes(label = country)
  ) +
  facet_grid(continent~.) +
  labs(title = "Change in GDP per capita and Life Expectancy between 1952-2007", 
       subtitle = "Question 5 - Plot 4 ",
       x = "GDP per capita ($)",
       y = "Life Expectancy (years)"
       )

describe_country %>%
  filter(continent == "Africa") %>% 
  ggplot(
    aes(
      x = gdpPercap_1952, 
      y = lifeExp_1952
      )
  ) +
  geom_segment(
    aes(
      xend = gdpPercap_2007, 
      yend = lifeExp_2007, 
      color = "Increases"
      ), 
    color = "red",
    arrow = arrow(length = unit(0.1,"cm"))
  ) +
   geom_segment(
     data = filter(weirdos, continent == "Africa"),
     mapping = aes(
      xend = gdpPercap_2007, 
      yend = lifeExp_2007, 
      color = "Decreases"
      ),
     color = "black",
     arrow = arrow(length = unit(0.1,"cm"))
  ) +
  geom_label_repel(
    data = filter(weirdos, continent == "Africa"),
    mapping = aes(label = country)
  ) +
  labs(title = "Change in GDP per capita and Life Expectancy in Africa between 1952-2007", 
       subtitle = "Question 5 - Plot 5 ",
       x = "GDP per capita ($)",
       y = "Life Expectancy (years)"
       )



```

***Observations***

* **Europe and Asia demonstrate the greates change in GDP per capita and Life Expectancy.**
* **In the Americas and Asia, only Kuwait, Haiti, and Nicaragua demonstrate a decrease in either GDP per capita or life expectancy between 1952-2007.**
* **In Africa, 11 countries demonstrate a decrease in either GDP per capita or life expectancy between 1952-2007: Zimbabwe, Swaziland, Comoros, Democractic Republic of Congo, Liberia, Madagascar, Niger, Centreal African REpublic, Djibouti, Sierra Leone, and Somalia.**
* **Most countries are "richer" and have a longer life expectancy between 1952-2007.**