---
title: "Final Challenge"
author: "ingridmathilde"
date: 2020-07-22
output:
  github_document:
    toc: true
---

*Purpose*: Explore Food Consumption vs CO2.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

#Setup & Data Wrangling

```{r setup}
library(tidyverse)
library(gapminder)
library(ggrepel)
```


```{r load all datasets}
food_consumption <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-18/food_consumption.csv')
gapminder_pop_total <- read_csv("./data/population_total.csv")
gapminder_geo <- read_csv("./data/countries_gapminder.csv")
```

```{r wrangle that data}
gapminder_pop_2014 <-
  gapminder_pop_total %>%
  select("country", "population" = "2014")

gapminder_geo_select <-
  gapminder_geo %>%
  select(
    "country" = "name", 
    "region" = "World bank region", 
    "income_grp" = "World bank, 4 income groups 2017"
  )

###All shamelessly copied from Angela's github###

## First, I'm fixing the USA
food_consumption_US_only <-
  food_consumption %>%
  filter(country == "USA") %>%
  select(-country) %>%
  mutate(country = "United States") %>%
  select(country, food_category, consumption, co2_emmission)

## Next, I'm fixing Hong Kong
food_consumption_HK_only <-
  food_consumption %>%
  filter(country == "Hong Kong SAR. China") %>%
  select(-country) %>%
  mutate(country = "Hong Kong, China") %>%
  select(country, food_category, consumption, co2_emmission)

## Next, I'm fixing Taiwan
food_consumption_TW_only <-
  food_consumption %>%
  filter(country == "Taiwan. ROC") %>%
  select(-country) %>%
  mutate(country = "Taiwan") %>%
  select(country, food_category, consumption, co2_emmission)

## Macedonia (Macedonia, FYR in gapminder)
food_consumption_Macedonia_only <-
  food_consumption %>%
  filter(country == "Macedonia") %>%
  select(-country) %>%
  mutate(country = "Macedonia, FYR") %>%
  select(country, food_category, consumption, co2_emmission)

## Congo, Rep.
food_consumption_CongoRep_only <-
  food_consumption %>%
  filter(country == "Congo") %>%
  select(-country) %>%
  mutate(country = "Congo, Rep.") %>%
  select(country, food_category, consumption, co2_emmission)


## Now, I'm binding them all together!
food_consumption_mod <-
  food_consumption %>%
  filter(
    country != "USA" &
    country != "Hong Kong SAR. China" &
    country != "Taiwan. ROC" &
    country != "Macedonia" &
    country != "Congo"
  ) %>%
  bind_rows(
    food_consumption_US_only,
    food_consumption_HK_only,
    food_consumption_TW_only,
    food_consumption_Macedonia_only,
    food_consumption_CongoRep_only
    )
df_food_pop <-
  food_consumption_mod %>%
  left_join(gapminder_pop_2014, by = "country") %>%
  left_join(gapminder_geo_select, by = "country") %>%
  group_by(country) %>%
  mutate(
    co2_food_country = sum(co2_emmission), 
    food_consumption_country = sum(consumption)
  ) %>%
  ungroup()

```


#Population Calculations

```{r step1 join}
df_food_pop_totals <- df_food_pop %>% 
  mutate(
    total_consumption_food = consumption*population,
    total_co2_food = co2_emmission*population
  )

df_food_pop_totals
```



***Question to answer here:* Who are the outliers for milk (including cheese) in terms of total consumption?**
```{r milk total consumption/co2}
total_co2_outliers <- df_food_pop_totals %>% 
  group_by(food_category) %>% 
  mutate(
    IQRange = IQR(total_co2_food, na.rm = TRUE),
    lower = quantile(total_co2_food, 0.25, na.rm = TRUE) - 1.5 * IQR(total_co2_food, na.rm = TRUE),
    upper = quantile(total_co2_food, 0.75, na.rm = TRUE) + 1.5 * IQR(total_co2_food, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  filter(upper < total_co2_food | total_co2_food < lower)

df_food_pop_totals %>%
  filter(food_category %in% c("Milk - inc. cheese")) %>% 
  ggplot(
    aes(
      x = total_co2_food,
      y = food_category
    )
  ) +
  geom_boxplot(outlier.color = "red") +
  geom_label_repel(
    data = filter(total_co2_outliers, food_category %in% c("Milk - inc. cheese")),
    mapping = aes(label = country)
  ) +
  labs(
    title = "Total Milk-Related CO2 emissions",
    x = "CO2 emissions per country",
    y = " "
  ) +
  scale_x_log10(labels = scales::label_number_si())


```

The outliers in total co2 emissions for milk including cheese are:

* India: largest population in the world and huge industry informed by [Hinduism](https://www.bloomberg.com/news/features/2018-01-23/india-s-sacred-cow-now-threatens-an-83-billion-dairy-industry).
* United States: despite [recent declines](https://www.theguardian.com/food/2020/jan/06/us-dairy-industry-suffering-americans-consume-less-milk), the total consumption is still high
* China: Despite [extremely high rates](https://www.foodsmatter.com/allergy_intolerance/cow's_milk_allergy_general/articles/dairy_intol_china.html#:~:text=Although%20there%20are%20no%20official,people%20to%20drink%20more%20milk.) of adult lactose intolerance, it is marketed as a [food of affluence](https://www.theguardian.com/environment/2019/mar/29/can-the-world-quench-chinas-bottomless-thirst-for-milk) and progress and is therefore consumed enormously.
* Pakistan: Pakistan is the sixth most populous country in the world and around half of households have [at least one milk-producing cow](http://www.fao.org/3/i0588e/I0588E07.htm)
* Brazil: huge population and [huge cow industry](https://www.cheesemarketnews.com/articlearch/passport/brazil.html)
* Russia: Big population I guess?
* Germany: 13th in [milk consumption per capita](https://en.wikipedia.org/wiki/List_of_countries_by_milk_consumption_per_capita)
* France: huge cheese producer
* United Kingdom: 
* Italy: 24th in [milk consumption per capita](https://en.wikipedia.org/wiki/List_of_countries_by_milk_consumption_per_capita)
* Turkey: 34th in [milk consumption per capita](https://en.wikipedia.org/wiki/List_of_countries_by_milk_consumption_per_capita)
* Mexico: 78th in [milk consumption per capita](https://en.wikipedia.org/wiki/List_of_countries_by_milk_consumption_per_capita)
* Argentina: 33th in [milk consumption per capita](https://en.wikipedia.org/wiki/List_of_countries_by_milk_consumption_per_capita)
* Japan: 109th in [milk consumption per capita](https://en.wikipedia.org/wiki/List_of_countries_by_milk_consumption_per_capita) and seems to be a recent addition to the [diet](https://lib.dr.iastate.edu/cgi/viewcontent.cgi?referer=https://www.google.com/&httpsredir=1&article=1420&context=card_workingpapers)
* Grenada: ???

```{r outliers examinination}

total_co2_outliers %>%
  filter(food_category %in% c("Milk - inc. cheese")) %>% 
  ggplot(
    aes(
      x = co2_emmission,
      y = reorder(country, co2_emmission)
    )
  ) +
  geom_col(
    aes(
      fill = region
    )
  ) +
  labs(
    title = "CO2 emissions due to milk per capita for \noutliers of total CO2 emissions by country",
    x = "CO2 emissions due to milk (including cheese) per capita",
    y = "Country"
  )


```

India and China have a really low consumption per capita. Most of the countries with higher per capita milk consumption are in Europe and Central Asia. 

Is there a relationship between milk consumption and population?

```{r milk consumption vs population}


df_food_pop_totals %>% 
  filter(food_category == "Milk - inc. cheese") %>% 
  ggplot() +
  geom_point(
    aes(
      x = population,
      y = co2_emmission
    )
  ) +   
  labs(
    title = "Milk-Related CO2 vs Population",
    x = "Population",
    y = "CO2 emissions per capita"
  ) +
  scale_x_log10(labels = scales::label_number_si())

```

Nope! No relationship there really.

```{r milk vs beef}
beef_v_milk <- df_food_pop_totals %>% 
  filter(food_category %in% c("Milk - inc. cheese", "Beef")) %>%
  select(
    country,
    region,
    income_grp,
    co2_food_country,
    food_category,
    co2_emmission,
    consumption
  ) %>% 
  pivot_wider(names_from = food_category, values_from = c(co2_emmission, consumption))
  
beef_v_milk

beef_v_milk %>% 
  ggplot() +
  geom_point(
    aes(
      x = co2_emmission_Beef,
      y = `co2_emmission_Milk - inc. cheese`,
      color = region
    )
  ) +   
  labs(
    title = "Milk vs Beef: CO2 Emissions",
    x = "Beef CO2 emissions per capita",
    y = "Milk CO2 emissions per capita"
  ) +
  scale_x_log10(labels = scales::label_number_si()) +
  scale_y_log10(labels = scales::label_number_si())

beef_v_milk %>% 
  ggplot() +
  geom_point(
    aes(
      x = consumption_Beef,
      y = `consumption_Milk - inc. cheese`,
      color = region
    )
  ) +   
  labs(
    title = "Milk vs Beef: Consumption",
    x = "Beef Consumptin per capita",
    y = "Milk Consumption per capita"
  ) +
  scale_x_log10(labels = scales::label_number_si()) +
  scale_y_log10(labels = scales::label_number_si())

beef_v_milk %>% 
  ggplot() +
  geom_point(
    aes(
      x = co2_emmission_Beef,
      y = `co2_emmission_Milk - inc. cheese`,
      color = income_grp
    )
  ) +   
  labs(
    title = "Milk vs Beef: CO2 Emissions",
    x = "Beef CO2 emissions per capita",
    y = "Milk CO2 emissions per capita"
  ) +
  scale_x_log10(labels = scales::label_number_si()) +
  scale_y_log10(labels = scales::label_number_si())


```

On a log-log scale, there is a linear relationship between beef and milk consumption/co2-related emissions. In general, higher income countries just consume more of both beef and milk - again, makes sense. Makes me wonder - between these income groups, is there a higher consumption of certain foods over others?

```{r most eaten foods}
df_percents <- df_food_pop_totals %>% 
  mutate(
    percent_diet = consumption/food_consumption_country,
    percent_co2 = co2_emmission/co2_food_country
  ) %>% 
  select(
    country,
    region,
    income_grp,
    food_category,
    percent_diet,
    percent_co2
  ) 

df_percents
```

```{r precent diet by region and income group}
df_percents %>%
  ggplot() +
  geom_boxplot(
    aes(
      x = percent_diet,
      y = food_category
    )
  ) +
  facet_wrap(.~region)


df_percents %>%
  ggplot() +
  geom_boxplot(
    aes(
      x = percent_diet,
      y = reorder(food_category, percent_diet, median)
    )
  ) +
  facet_wrap(.~income_grp)

  
```

Top 3 by percent diet by region:
 
* East Asia & Pacific: Rice, Wheat, Milk & Cheese
* Europe & Central Asia: Milk & Cheese, Wheat, Pork
* Latin America & Caribbean: Milk & Cheese, Wheat, Poultry
* Middle East & North Africa: Wheat, Milk & Cheese, Rice
* North America: Milk & Cheese, Wheat, Poultry
* South Asia: Rice, Milk & Cheese, Wheat
* Sub-Saharan Africa: Wheat, Milk & Cheese, Rice
* NA: Milk & Cheese, Wheat, Poultry

Top 3 percent diet by income group:

* High income: Milk & Cheese, Wheat, Pork
* Low income: Rice, Wheat, Milk & Cheese
* Lower middle income: Wheat, Milk & Cheese, Rice
* Upper middle income: Milk & Cheese, Wheat, Poultry
* NA: Milk & Cheese, Wheat, Poultry

```{r percent co2 emissions}
df_percents %>%
  ggplot() +
  geom_boxplot(
    aes(
      x = percent_co2,
      y = food_category
    )
  ) +
  facet_wrap(.~region)
```


Top Food Emissions by region (I predict beef...):

* East Asia & Pacific: Beef
* Europe & Central Asia: Beef
* Latin America & Caribbean: Beef
* Middle East & North Africa: Beef
* North America: Beef
* South Asia: Rice (that must be a lot of rice...)
* Sub-Saharan Africa: Beef
* NA: Beef

That was sort of boring.Beef was the top when mapping co2 emissions vs income group.

```{r per person milk consumption/co2}
total_co2_outliers <- df_food_pop_totals %>% 
  group_by(food_category) %>% 
  mutate(
    IQRange = IQR(co2_emmission, na.rm = TRUE),
    lower = quantile(co2_emmission, 0.25, na.rm = TRUE) - 1.5 * IQR(co2_emmission, na.rm = TRUE),
    upper = quantile(co2_emmission, 0.75, na.rm = TRUE) + 1.5 * IQR(co2_emmission, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  filter(upper < co2_emmission)

df_food_pop_totals %>%
  filter(food_category %in% c("Milk - inc. cheese")) %>% 
  ggplot(
    aes(
      x = co2_emmission,
      y = food_category
    )
  ) +
  geom_boxplot(outlier.color = "red") +
  geom_label_repel(
    data = filter(total_co2_outliers, food_category %in% c("Milk - inc. cheese")),
    mapping = aes(label = country)
  ) +
  labs(
    title = "Consumption by food category"
  )

```

Finland is the only outlier here with well above the median milk consumption per capita. But this is sort of boring because we all know Northern Europe drinks a lot of milk...

```{r total food consumption}
group_width <- 50

consumption_groups <- df_food_pop_totals %>% 
  select(
    country,
    region,
    income_grp,
    population,
    food_consumption_country,
    co2_food_country
  ) %>% 
  distinct() %>% 
  mutate(
    food_consumption_group = cut_width(food_consumption_country, group_width)
  )

consumption_group_outliers <- consumption_groups %>% 
  group_by(food_consumption_group) %>%
  mutate(
    IQRange = IQR(co2_food_country, na.rm = TRUE),
    lower = quantile(co2_food_country, 0.25, na.rm = TRUE) - 1.5 * IQR(co2_food_country, na.rm = TRUE),
    upper = quantile(co2_food_country, 0.75, na.rm = TRUE) + 1.5 * IQR(co2_food_country, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  filter(upper < co2_food_country | co2_food_country < lower)


df_food_pop_totals %>% 
  ggplot() +
  geom_point(
    aes(
      x = food_consumption_country,
      y = co2_food_country,
      color = income_grp
    )
  ) +
  labs(
    title = "CO2 emissions per capita as a function of Consumption per capita",
    x = "Total Food Consumption per capita (kg/person/year)",
    y = "Total Food-Related CO2 Emissions per capita (kg/per/year)",
    color = "Income Group"
  )
  

df_food_pop_totals %>% 
  ggplot(
    aes(
      x = food_consumption_country,
      y = co2_food_country,
      group = cut_width(food_consumption_country, group_width)
    )
  ) +
  geom_boxplot(
    outlier.color = "red" 
  ) +
  geom_label_repel(
    data = consumption_group_outliers,
    mapping = aes(
      label = country,
      color = income_grp
    )
  ) +
  labs(
    title = "CO2 emissions per capita as a function of Consumption per capita",
    x = "Total Food Consumption per capita\n (kg/person/year)",
    y = "Total Food-Related CO2 Emissions per capita\n (kg/per/year)",
    color = "Income Group"
  )
```

So no shit, but food consumption is directly correlated to CO2 emissions... cool. When broken down by consumption, there are a few outliers aka places where they consume similar total food (kg/person/year) but do not produce as much CO2 (kg/person/year) - I predict this is dictated by the breakdown of the foods they eat. What jumps out to me as interesting is the contrast between the United States and Maldives - they are in the same range for total food consumption but have vastly different CO2 emissions - how different are they in terms of dietary breakdown? I predict that the United States eats way more beef.

```{r USA and Maldives percents}

df_percents %>% 
  filter(country %in% c("United States", "Maldives")) %>% 
  select(
    country,
    food_category,
    percent_diet
  ) %>% 
  ggplot() +
  geom_col(
    aes(
      x = country,
      y = percent_diet,
      fill = food_category
    )
  )

df_percents %>% 
  filter(country %in% c("United States", "Maldives")) %>% 
  select(
    country,
    food_category,
    percent_co2,
    percent_diet
  ) %>% 
  ggplot() +
  geom_col(
    aes(
      x = country,
      y = percent_co2,
      fill = food_category
    )
  )

```

Overall, the United States eats more red meat while Maldives eats more fish and rice. These are the main contributors to  having vastly higher CO2 emmissions.

```{r USA and Maldives totals}

df_food_pop_totals %>% 
  filter(country %in% c("United States", "Maldives")) %>% 
  select(
    country,
    food_category,
    consumption
  ) %>% 
  ggplot() +
  geom_col(
    aes(
      x = country,
      y = consumption,
      fill = food_category
    )
  )

df_food_pop_totals %>% 
  filter(country %in% c("United States", "Maldives")) %>% 
  select(
    country,
    food_category,
    co2_emmission
  ) %>% 
  ggplot() +
  geom_col(
    aes(
      x = country,
      y = co2_emmission,
      fill = food_category
    )
  )

```

Because the USA consumes more beef, it has much higher emissions. 

```{r BRIC-U}
df_BRICU_percents <- df_percents %>% 
  filter(country %in% c("United States", "Brazil", "Russia", "India", "China")) %>% 
  select(
    country,
    food_category,
    percent_diet
  ) %>%
  ggplot() +
  geom_col(
    aes(
      x = country,
      y = percent_diet,
      fill = food_category
    )
  )

df_BRICU_percents
```

```{r high income}
df_high_income_percent <- df_percents %>% 
  filter(income_grp == "High income") %>% 
  filter()

df_high_income_totals <- df_food_pop_totals  %>% 
  filter(income_grp == "High income")

df_high_income_percent
df_high_income_totals

group_width <- 50

high_income_consumption_groups <- df_high_income_totals %>% 
  select(
    country,
    region,
    income_grp,
    population,
    food_consumption_country,
    co2_food_country
  ) %>% 
  distinct() %>% 
  mutate(
    food_consumption_group = cut_width(food_consumption_country, group_width)
  )

high_income_consumption_group_outliers <- high_income_consumption_groups %>% 
  group_by(food_consumption_group) %>%
  mutate(
    IQRange = IQR(co2_food_country, na.rm = TRUE),
    lower = quantile(co2_food_country, 0.25, na.rm = TRUE) - 1.5 * IQR(co2_food_country, na.rm = TRUE),
    upper = quantile(co2_food_country, 0.75, na.rm = TRUE) + 1.5 * IQR(co2_food_country, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  filter(upper < co2_food_country | co2_food_country < lower)

high_income_consumption_groups %>% 
  ggplot(
    aes(
      x = food_consumption_country,
      y = co2_food_country,
      group = cut_width(food_consumption_country, group_width)
    )
  ) +
  geom_boxplot(
    outlier.color = "red" 
  ) +
  geom_label_repel(
    data = high_income_consumption_group_outliers,
    mapping = aes(
      label = country,
      color = income_grp
    )
  ) +
  labs(
    title = "CO2 emissions per capita as a function of Consumption per capita",
    x = "Total Food Consumption per capita\n (kg/person/year)",
    y = "Total Food-Related CO2 Emissions per capita\n (kg/per/year)",
    color = "Income Group"
  )

high_income_consumption_groups %>% 
  ggplot(
    aes(
      x = food_consumption_country,
      y = co2_food_country,
      group = cut_width(food_consumption_country, group_width)
    )
  ) +
  geom_point() +
  labs(
    title = "CO2 emissions per capita as a function of Consumption per capita",
    x = "Total Food Consumption per capita\n (kg/person/year)",
    y = "Total Food-Related CO2 Emissions per capita\n (kg/per/year)",
    color = "Income Group"
  )

high_income_consumption_groups %>% 
  filter(food_consumption_country >500 & co2_food_country < 1000)

high_income_consumption_groups %>% 
  filter(food_consumption_country >400 & co2_food_country < 1000)

```

```{r final data set}
df_food_all <-
  food_consumption_mod %>%
  left_join(gapminder_pop_2014, by = "country") %>%
  left_join(gapminder_geo_select, by = "country") %>%
  filter(food_category %in% c("Beef", "Fish", "Lamb & Goat", "Pork", "Poultry", "Eggs", "Milk - inc. cheese")) %>% 
  group_by(country) %>%
  mutate(
    co2_food_country = sum(co2_emmission), 
    food_consumption_country = sum(consumption)
  ) %>%
  ungroup() %>% 
  mutate(
    percent_diet = consumption/food_consumption_country,
    percent_co2 = co2_emmission/co2_food_country
  )

df_food_high_income <- df_food_all %>% 
  filter(income_grp == "High income")
```

```{r and some final plots}
group_width <- 50

high_income_groups <- df_food_high_income %>% 
  select(
    country,
    region,
    income_grp,
    population,
    food_consumption_country,
    co2_food_country
  ) %>% 
  distinct() %>% 
  mutate(
    food_consumption_group = cut_width(food_consumption_country, group_width)
  )

high_income_group_outliers <- high_income_groups %>% 
  group_by(food_consumption_group) %>%
  mutate(
    IQRange = IQR(co2_food_country, na.rm = TRUE),
    lower = quantile(co2_food_country, 0.25, na.rm = TRUE) - 1.5 * IQR(co2_food_country, na.rm = TRUE),
    upper = quantile(co2_food_country, 0.75, na.rm = TRUE) + 1.5 * IQR(co2_food_country, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  filter(upper < co2_food_country | co2_food_country < lower)

high_income_beef <- df_food_high_income %>% 
  filter(food_category == "Beef")

high_income_groups %>% 
  ggplot(
    aes(
      x = food_consumption_country,
      y = co2_food_country,
      group = cut_width(food_consumption_country, group_width)
    )
  ) +
  geom_boxplot(
    outlier.color = "red" 
  ) +
  geom_label_repel(
    data = high_income_group_outliers,
    mapping = aes(
      label = country,
      color = income_grp
    )
  ) +
  labs(
    title = "CO2 emissions per capita as a function of Consumption per capita",
    x = "Total Food Consumption per capita\n (kg/person/year)",
    y = "Total Food-Related CO2 Emissions per capita\n (kg/per/year)",
    color = "Income Group"
  )

high_income_groups %>% 
  ggplot(
    aes(
      x = food_consumption_country,
      y = co2_food_country,
    )
  ) +
  geom_point(
    aes(
      alpha = high_income_beef$percent_diet
    )
  ) +
  labs(
    title = "CO2 emissions per capita as a function of Consumption per capita",
    x = "Total Food Consumption per capita\n (kg/person/year)",
    y = "Total Food-Related CO2 Emissions per capita\n (kg/per/year)",
    color = "Income Group"
  )

```