---
title: "Final Challenge"
author: "ingridmathilde"
date: 2020-08-07
output:
  github_document:
    toc: true
---

*Purpose*: Explore Food Consumption vs CO2.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

#Setup & Data Wrangling

```{r setup}
library(tidyverse)
library(gapminder)
library(ggrepel)
```


```{r load all datasets}
food_consumption <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-18/food_consumption.csv')
gapminder_pop_total <- read_csv("./data/population_total.csv")
gapminder_geo <- read_csv("./data/countries_gapminder.csv")
```

```{r wrangle that data}
gapminder_pop_2014 <-
  gapminder_pop_total %>%
  select("country", "population" = "2014")

gapminder_geo_select <-
  gapminder_geo %>%
  select(
    "country" = "name", 
    "region" = "World bank region", 
    "income_grp" = "World bank, 4 income groups 2017"
  )

###All shamelessly copied from Angela's github###

## First, I'm fixing the USA
food_consumption_US_only <-
  food_consumption %>%
  filter(country == "USA") %>%
  select(-country) %>%
  mutate(country = "United States") %>%
  select(country, food_category, consumption, co2_emmission)

## Next, I'm fixing Hong Kong
food_consumption_HK_only <-
  food_consumption %>%
  filter(country == "Hong Kong SAR. China") %>%
  select(-country) %>%
  mutate(country = "Hong Kong, China") %>%
  select(country, food_category, consumption, co2_emmission)

## Next, I'm fixing Taiwan
food_consumption_TW_only <-
  food_consumption %>%
  filter(country == "Taiwan. ROC") %>%
  select(-country) %>%
  mutate(country = "Taiwan") %>%
  select(country, food_category, consumption, co2_emmission)

## Macedonia (Macedonia, FYR in gapminder)
food_consumption_Macedonia_only <-
  food_consumption %>%
  filter(country == "Macedonia") %>%
  select(-country) %>%
  mutate(country = "Macedonia, FYR") %>%
  select(country, food_category, consumption, co2_emmission)

## Congo, Rep.
food_consumption_CongoRep_only <-
  food_consumption %>%
  filter(country == "Congo") %>%
  select(-country) %>%
  mutate(country = "Congo, Rep.") %>%
  select(country, food_category, consumption, co2_emmission)


## Now, I'm binding them all together!
food_consumption_mod <-
  food_consumption %>%
  filter(
    country != "USA" &
    country != "Hong Kong SAR. China" &
    country != "Taiwan. ROC" &
    country != "Macedonia" &
    country != "Congo"
  ) %>%
  bind_rows(
    food_consumption_US_only,
    food_consumption_HK_only,
    food_consumption_TW_only,
    food_consumption_Macedonia_only,
    food_consumption_CongoRep_only
    )

```


```{r final data set}
df_food_all <-
  food_consumption_mod %>%
  left_join(gapminder_pop_2014, by = "country") %>%
  left_join(gapminder_geo_select, by = "country") %>%
  filter(food_category %in% c("Beef", "Fish", "Lamb & Goat", "Pork", "Poultry", "Eggs", "Milk - inc. cheese")) %>% 
  group_by(country) %>%
  mutate(
    co2_food_country = sum(co2_emmission), 
    food_consumption_country = sum(consumption)
  ) %>%
  ungroup() %>% 
  mutate(
    percent_diet = consumption/food_consumption_country,
    percent_co2 = co2_emmission/co2_food_country
  )

df_food_high_income <- df_food_all %>% 
  filter(income_grp == "High income")
```

```{r and some final plots}
group_width <- 50

high_income_groups <- df_food_high_income %>% 
  select(
    country,
    region,
    income_grp,
    population,
    food_consumption_country,
    co2_food_country
  ) %>% 
  distinct() %>% 
  mutate(
    food_consumption_group = cut_width(food_consumption_country, group_width)
  )

high_income_group_outliers <- high_income_groups %>% 
  group_by(food_consumption_group) %>%
  mutate(
    IQRange = IQR(co2_food_country, na.rm = TRUE),
    lower = quantile(co2_food_country, 0.25, na.rm = TRUE) - 1.5 * IQR(co2_food_country, na.rm = TRUE),
    upper = quantile(co2_food_country, 0.75, na.rm = TRUE) + 1.5 * IQR(co2_food_country, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  filter(upper < co2_food_country | co2_food_country < lower)

high_income_beef <- df_food_high_income %>% 
  filter(food_category == "Beef")

high_income_groups %>% 
  ggplot(
    aes(
      x = food_consumption_country,
      y = co2_food_country,
      group = cut_width(food_consumption_country, group_width)
    )
  ) +
  geom_boxplot(
    outlier.color = "red" 
  ) +
  geom_label_repel(
    data = high_income_group_outliers,
    mapping = aes(
      label = country,
      color = income_grp
    )
  ) +
  labs(
    title = "CO2 emissions per capita as a function of Consumption per capita",
    x = "Total Food Consumption per capita\n (kg/person/year)",
    y = "Total Food-Related CO2 Emissions per capita\n (kg/per/year)",
    color = "Income Group"
  )

high_income_groups %>% 
  ggplot(
    aes(
      x = food_consumption_country,
      y = co2_food_country,
    )
  ) +
  geom_point(
    aes(
      alpha = high_income_beef$percent_diet
    )
  ) +
  labs(
    title = "CO2 emissions per capita as a function of Consumption per capita",
    x = "Total Food Consumption per capita\n (kg/person/year)",
    y = "Total Food-Related CO2 Emissions per capita\n (kg/per/year)",
    color = "Income Group"
  )

```