---
title: "Final Challenge"
author: "ingridmathilde"
date: 2020-08-07
output:
  github_document:
    toc: true
---

*Purpose*: Explore Food Consumption vs CO2.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

#Setup & Data Wrangling

```{r setup}
library(tidyverse)
library(gapminder)
library(ggrepel)
library(wesanderson)
```


```{r load all datasets}
food_consumption <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-18/food_consumption.csv')
gapminder_pop_total <- read_csv("./data/population_total.csv")
gapminder_geo <- read_csv("./data/countries_gapminder.csv")
```

```{r wrangle that data}
gapminder_pop_2014 <-
  gapminder_pop_total %>%
  select("country", "population" = "2014")

gapminder_geo_select <-
  gapminder_geo %>%
  select(
    "country" = "name", 
    "region" = "World bank region", 
    "income_grp" = "World bank, 4 income groups 2017"
  )

###All shamelessly copied from Angela's github###

## First, I'm fixing the USA
food_consumption_US_only <-
  food_consumption %>%
  filter(country == "USA") %>%
  select(-country) %>%
  mutate(country = "United States") %>%
  select(country, food_category, consumption, co2_emmission)

## Next, I'm fixing Hong Kong
food_consumption_HK_only <-
  food_consumption %>%
  filter(country == "Hong Kong SAR. China") %>%
  select(-country) %>%
  mutate(country = "Hong Kong, China") %>%
  select(country, food_category, consumption, co2_emmission)

## Next, I'm fixing Taiwan
food_consumption_TW_only <-
  food_consumption %>%
  filter(country == "Taiwan. ROC") %>%
  select(-country) %>%
  mutate(country = "Taiwan") %>%
  select(country, food_category, consumption, co2_emmission)

## Macedonia (Macedonia, FYR in gapminder)
food_consumption_Macedonia_only <-
  food_consumption %>%
  filter(country == "Macedonia") %>%
  select(-country) %>%
  mutate(country = "Macedonia, FYR") %>%
  select(country, food_category, consumption, co2_emmission)

## Congo, Rep.
food_consumption_CongoRep_only <-
  food_consumption %>%
  filter(country == "Congo") %>%
  select(-country) %>%
  mutate(country = "Congo, Rep.") %>%
  select(country, food_category, consumption, co2_emmission)


## Now, I'm binding them all together!
food_consumption_mod <-
  food_consumption %>%
  filter(
    country != "USA" &
    country != "Hong Kong SAR. China" &
    country != "Taiwan. ROC" &
    country != "Macedonia" &
    country != "Congo"
  ) %>%
  bind_rows(
    food_consumption_US_only,
    food_consumption_HK_only,
    food_consumption_TW_only,
    food_consumption_Macedonia_only,
    food_consumption_CongoRep_only
    )

```


```{r final data set}
df_food_all <-
  food_consumption_mod %>%
  left_join(gapminder_pop_2014, by = "country") %>%
  left_join(gapminder_geo_select, by = "country") %>%
  filter(food_category %in% c("Beef", "Fish", "Lamb & Goat", "Pork", "Poultry", "Eggs", "Milk - inc. cheese")) %>% 
  group_by(country) %>%
  mutate(
    co2_food_country = sum(co2_emmission), 
    food_consumption_country = sum(consumption)
  ) %>%
  ungroup() %>% 
  mutate(
    percent_diet = consumption/food_consumption_country,
    percent_co2 = co2_emmission/co2_food_country
  )

df_food_high_income <- df_food_all %>% 
  filter(income_grp == "High income")
```

```{r and some final plots}
group_width <- 100

high_income_groups <- df_food_high_income %>% 
  select(
    country,
    region,
    income_grp,
    population,
    food_consumption_country,
    co2_food_country
  ) %>% 
  distinct() %>% 
  mutate(
    food_consumption_group = cut_width(food_consumption_country, group_width)
  )

high_income_group_outliers <- high_income_groups %>% 
  group_by(food_consumption_group) %>%
  mutate(
    IQRange = IQR(co2_food_country, na.rm = TRUE),
    lower = quantile(co2_food_country, 0.25, na.rm = TRUE) - 1.5 * IQR(co2_food_country, na.rm = TRUE),
    upper = quantile(co2_food_country, 0.75, na.rm = TRUE) + 1.5 * IQR(co2_food_country, na.rm = TRUE)
  ) %>% 
  ungroup() %>% 
  filter(upper < co2_food_country | co2_food_country < lower)

high_income_groups %>% 
  ggplot(
    aes(
      x = food_consumption_group,
      y = co2_food_country
    )
  ) +
  geom_boxplot() +
  labs(
     title = "Animal and Aminal Byproduct-Related CO2 emissions \ncorrelates with Consumption of said Foods",
    subtitle = "For High Income Countries",
    x = "Consumption per capita\n (kg/person/year)",
    y = "CO2 Emissions per capita\n (kg/per/year)",
    color = "Beef as Percent\n of Animal Diet (%)"
  ) +
  theme(plot.title = element_text(face = "bold")) +
  scale_color_manual(values=wes_palette(n=4, name="Darjeeling1")) 

```

```{r beef eaters}
high_income_beef <- df_food_high_income %>% 
  filter(food_category == "Beef") %>% 
  mutate(
    beef_group = cut_width(percent_diet*100,4, boundary = 0)
  )

high_income_groups %>% 
  ggplot(
    aes(
      x = food_consumption_country,
      y = co2_food_country,
    )
  ) +
  geom_point(
    aes(
      color = high_income_beef$beef_group
    )
  ) +
  labs(
    title = "Countries that consume more beef tend to produce\n more CO2 emissions",
    subtitle = "For High Income Countries",
    x = "Animal (and Byproduct) Consumption per capita\n (kg/person/year)",
    y = "Animal (and Byproduct) Related-CO2 Emissions\n per capita\n (kg/per/year)",
    color = "Beef as Percent\n of Animal Diet (%)"
  ) +
  theme(plot.title = element_text(face = "bold")) +
  scale_color_manual(values=wes_palette(n=4, name="Darjeeling1")) 
```

```{r countries to emulate}
df_companion_countries <- df_food_high_income %>% 
  mutate(
    food_consumption_group = cut_width(food_consumption_country, group_width)
  ) %>% 
  filter(food_consumption_group == "(350,450]")

df_companion_countries

df_companion_countries %>% 
  ggplot() +
  geom_col(
    aes(
      y = reorder(country, co2_food_country),
      x = percent_diet,
      fill = food_category
    )
  ) +
  labs(
    title = "Milk and Cheese is a major food group!",
    subtitle = "For High Income Countries",
    x = "Percentage of Diet (%)",
    y = "Country",
    color = "Animal or Animal Byproduct"
  ) +
  theme(plot.title = element_text(face = "bold")) +
  scale_color_manual(values=wes_palette(n=4, name="Darjeeling1")) 

df_companion_countries %>% 
  ggplot() +
  geom_col(
    aes(
      y = reorder(country, co2_food_country),
      x = co2_emmission,
      fill = food_category
    )
  )+
  labs(
    title = "Beef, Lamb, and Goat are major CO2 emittors",
    subtitle = "For High Income Countries",
    x = "CO2 Emissions (kg/person/year)",
    y = "Country",
    color = "Animal or Animal Byproduct"
  ) +
  theme(plot.title = element_text(face = "bold")) +
  scale_color_manual(values=wes_palette(n=4, name="Darjeeling1")) 
```
